Building the Million Transcriptome Ocular Meta-Atlas

### 2020-12-09 manuscript_00.Rmd
text updates

## The meta-atlas

We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas. As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc). The benchmarking system we developed uses a wide variety of metrics and we make available via the R package scPOP (single-cell Pick Optimal Parameters). The scEiaD will be of utility to two communities: the ocular community who can both search scEiaD for gene expression across many dimensions (e.g. cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and automatic cell labelling. Also the computational community  can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment. As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base and propose general guidelines to optimally create custom meta-atlases.

@@

scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implmenets the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms. The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better). NMI and ARI were used to assess the consistency of cell type to cluster assignment (where 1 is perfect correspondence between cluster and cell type).

@@

## scPOP reveals different priorities across the different batch integration methods

On one extreme we have combat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a). The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each study is a distinct in the 2D UMAP space (`r fig_cap(name = 'figBenchmark', display = 'cite')`b). In our scEiaD dataset we see that like combat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity. Scanorama, fastMNN, and bbknn more prioritize cleanly separating the clusters. With our scEiaD meta-atlas insct, scArches, and magic do not perform particularly well in batch mixing or cluster purity.

@@

https://www.frontiersin.org/articles/10.3389/fgene.2020.00041/full
Another substantial area of uncertainty is expression normalization. The approach (which we call "standard") that Seurat and anndata use by default is to, per cell, divide the counts by the sum counts for the cell, multiply by a scaling factor, then log transform. This helps make the count distribution more normal, which is an assumption that many algorithms require. In contrast, the scran normalization method groups cells into pools and normalizes across the pool summed counts instead of the individual cell counts. We also use the square root (sqrt) normalization which replaces the log transformation with a square root. Library size (libSize) normalization omits the sqrt or log transformation. Finally some methods like scVI directly use the raw counts data for modeling the data.

We see in the across the sumZScale metric, which we define as sumZScale = zscale(LISI batch) + zscale(LISI cluster) + zscale(LISI cell type) + zscale(silhouette batch) + zscale(silhouette cluster) + zscale(silhouette cell type) + zscale(NMI) + zscale(ARI) that scVI overall performs best. As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c). We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods. We also see that changing the number of latent dimensions (8 or 30) can occassionally dramatically change performance.

@@

To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors. Furthermore we used a recent advance in scVI capability that allows one to build a reference atlas with a subset of cells and project the remaining cells onto it - we qualitatively saw that the human only integration had the most reliable performance across the key parameters. We thus built a human scVI model and projected the mouse and macaque data onto it. We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters. Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data. We found the optimal parameters to be 5000 HVG, 10 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by PARC (`r fig_cap(name = 'figBenchmark', display = 'cite')`d). We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.1 and 500 nearest neighbors.

## xgboost ML model built to label unknown cell types

@@

## ML cell type predictor has high accuracy across most cell types

@@

## Certain retina cell types have more stable community consensus IDs

With the wide diversity of the scEiaD we can get a sense on how consistent the retina community is at labelling the major cell types of the retina. By calculating the AUC against each individual cell type and study, we can see which cell types are consistently labeled correctly across most (`r fig_cap(name = 'supFig_PR_cap', display = 'cite')`). Most studies show very similar performance with our ML model. The study with the lowest correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.

@@

## The scVI-based scEiaD UMAP projection blends batches while separating cell types
The 2D UMAP projection of the scVI-calculated batch corrected 10 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types. We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization. The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space. The amacrine, muller glia, horizontal, and retinal ganglia are in distinct spaces.

>>>>

## The projectable meta-atlas

We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas. As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc). The benchmarking system we developed uses a wide variety of metrics that we make easily available via the R package scPOP (single-cell Pick Optimal Parameters). The scEiaD will be of utility to two communities: the ocular community who can both search scEiaD for gene expression across many dimensions (e.g. cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and automatic cell labelling. Also the computational community  can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment. As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base and propose general guidelines to optimally create custom meta-atlases.

@@

scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implemnets the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms. The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better). NMI and ARI were used to assess the consistency of cell type to cluster assignment (where 1 is perfect correspondence between cluster and cell type).

@@

<!-- ## scPOP reveals different priorities across the different batch integration methods -->

On one extreme we have combat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a). The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each nearly study is has a distinct space in the 2D (`r fig_cap(name = 'figBenchmark', display = 'cite')`b). In our scEiaD dataset we see that like combat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity. Scanorama, fastMNN, and bbknn more prioritize cleanly separating the clusters. With our scEiaD meta-atlas insct, scArches, and magic do not perform particularly well in batch mixing or cluster purity.

@@

<!-- https://www.frontiersin.org/articles/10.3389/fgene.2020.00041/full -->
There are several normalization approaches that have been used or published. The "standard" approach that the popular analysis packages Seurat and anndata use by default is to, per cell, divide the counts by the sum counts for the cell, multiply by a scaling factor, then log transform. This helps make the count distribution more normal, which is an assumption that many algorithms require. In contrast, the scran normalization method groups cells into pools and normalizes across the pool summed counts instead of the individual cell counts. We also use the square root (sqrt) normalization which replaces the log transformation with a square root. Library size (libSize) normalization omits the sqrt or log transformation. Finally some methods like scVI directly use the raw counts data for modeling the data.

We see in the across the sumZScale metric, which we define as sumZScale = zscale(LISI batch) + zscale(LISI cluster) + zscale(LISI cell type) + zscale(silhouette batch) + zscale(silhouette cluster) + zscale(silhouette cell type) + zscale(NMI) + zscale(ARI) that scVI overall performs best. As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c). We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods. We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance. These results demonstrate the importance of assessing performance in a rigorous manner aceoss many parameters.

@@

To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors. Furthermore we used a recent advance in scVI capability that allows one to build a reference model and project the new cells onto it. We thus built a human-only scVI model and then projected the mouse and macaque data onto it. We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters. Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data. We found the optimal parameters to be 5000 HVG, 10 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by PARC (`r fig_cap(name = 'figBenchmark', display = 'cite')`d). We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.1 and 500 nearest neighbors.

## High accuracy xgboost ML model built to label unknown cell types

@@

<!-- ## ML cell type predictor has high accuracy across most cell types -->

@@

<!-- ## Certain retina cell types have more stable community consensus IDs -->

<!-- With the wide diversity of the scEiaD we can get a sense on how consistent the retina community is at labelling the major cell types of the retina. By calculating the AUC against each individual cell type and study, we can see which cell types are consistently labeled correctly across most (`r fig_cap(name = 'supFig_PR_cap', display = 'cite')`). Most studies show very similar performance with our ML model. The study with the lowest correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.   -->

@@

## The scVI-based scEiaD UMAP projection blends batches while separating cell types

The 2D UMAP projection of the scVI-calculated batch corrected 10 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types. We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization. The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space. The amacrine, muller glia, horizontal, and retinal ganglia are in distinct spaces.
